Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-09-01 manu-draft.Rmd
started discussion text

<!-- Seagrass: note decline of HW and correlations in the matrix, these make sense.  -->

@@

* Explain results - what's up with inorganic nutrients?  Likely driving phyto response but concentrations are not a good indicator.  Uptake is rapid...
* Explain correlations/PCA - area 3 distinct from area 1 (high salinity/turtle grass dominated, likely just a natural grouping)
* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
* Potential long-term impacts TBD
* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
* Risk of decline (IRL ex.), regression of past progress

>>>>

@@

Several of the correlations in Figure \@ref(fig:allcor) can be explained through mechanistic relationshps among water quality, macroalgae, and seagrasses.  Conversely, many of the correlations may be spurious and due to collinearity among variables that is likely not causative.  Regarding the nutrient management paradigm for Tampa Bay, the linear association between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations and loadings from the PCA in Figure \@ref(fig:allpca). An increase in nitrogen was associated with an increase in chlorophyll, which in turn was associated with a decrease in water clarity.  Temporally, these changes were observed at different times and by different species of phytoplankton.  The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K. brevis* in June and July.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa.  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen already utilized by different taxa, including macroalgae.

From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability that may negatively affect seagrass growth.  However, changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous.  Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.  Spatially, area 3 closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay had higher frequency occurrence of turtle grass and lower frequency of shoal grass.  This reflects a natural partitioning of the species along a salinity gradient (as shown in the PCA in Figure \@ref(fig:allpca)) and not any association with effluent from Piney Point.  Correlations among the macroalgae and seagrasses can also be explained by spatial differences among species along a depth gradient. Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.  Seagrasses are clonal organisms that grow together and species are usually not randomly dispersed in the same patch.  Shoal grass and turtle grass typically grow at shallower depths than manatee grass and are often exposed during low tides.  As an adaptive mechanism to varying water levels, blades of shoal grass and turtle grass can "lay down" during low tide to minimize dessication stress.  Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species is better suited at deeper depths that are minimally affected by varying tidal height.  As such, the correlation analysis can be explained by spatial partitioning of seagrasses along depth gradients due to adaptive characteristics of each species.  These associations may also reflect natural variation in water clarity along the longitidinal axis of Tampa Bay.  Shoal grass can occur in more turbid waters that increase towards the northern portion of Tampa Bay, which are also shallower.

Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.  This was expected given both the documented changes from past releases from Piney Point and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97]. In Tampa Bay, red macroalgae groups (e.g., *Gracilaria sp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva sp.*, *Caulerpa sp.*) and seasonally occur earlier in the growing season.  The dominance of the red groups early in the period of record followed by an increase in *Ulva sp.* likely reflects a natural phenology in Tampa Bay.  However past spill events were associated with an increase in both red and green macroalgae in smaller subembayments of Tampa Bay, prompting manual removal to prevent conditions associated with hypoxia from decomposition of macroalgal biomass [@Switzer11]. Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites as a result of nutrient cycling stimulated by wastewater release. A clear difference in the macroalgae community in 2021 compared to past years was a high abundance of cyanobacteria macroalgae (i.e., Lyngbya sp.).  Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Lyngbya sp.* in Anna Maria Sound and near Port Manatee was observed in 2021. The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.  *Lyngbya sp* were commonly observed at depth and entangled in blades of manatee grass (positive correlation in Figure \@ref(fig:allcor)), whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.

The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.  Although the seagrass community did not change dramatically across the study period, long-term impacts may be observed.  Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that peristed over a year and half.  These changes may have resulted from a reduction in water clarity from heavy concentrations of *K. brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.    Annually, seagrass coverage in Tampa Bay has declined... what other factors? PP is an affront...

* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
* Potential long-term impacts TBD
* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
* Risk of decline (IRL ex.), regression of past progress
